---
title: "<center>**Introduction to working with vegan package in R**</center>"
author: "<center>Wyclife Agumba Oluoch (wyclifeoluoch@gmail.com) </center>"
date: "<center>`r Sys.time()`</center>"
bibliography: 
  - bib/packages.bib
nocite: '@*'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: darkly
    #highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
packages <- c("base",
              'knitr',
              'rmarkdown',
              'prettydoc',
              'vegan',
              'ggvegan',
              'ggpubr')
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == FALSE)){
  install.packages(packages[!installed_packages])
}
lapply(packages, library, character.only = TRUE) |> invisible()
```

```{r write_bib, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
knitr::write_bib(c(
  .packages(), packages
), 'bib/packages.bib')
```

# Background

I explore `vegan`. `vegan` is an R pack that features in day to day working of a community ecologist. It would be great to know a few cases where the package comes in handy.

# Load data

We load some inbuilt data, BCI, that come with the `vegan` package.

```{r}
data("BCI")
```

Seems there are a lot of data here, we can check the dimensions of the data.frame using the `dim` function from base `R` package.

```{r}
dim(BCI)
```

Another way of getting the number of species in the data-set is to use `specnumber()` function in `vegan` package.

```{r}
specnumber(BCI)
```

It is possible to estimate the number of species in the pool using `specpool()` function in `R`. So, actually there are 225 species, but the estimator gives values from 236 to 247 with some standard errors.

```{r}
specpool(BCI, smallsample = TRUE)
```

# Calculating selected diversity indices

Here is a data-set of 50 plots indicating number of individuals of species of trees counted in each plot. Typical ecological survey work.

So, we can go ahead and calculate species diversity in the 50 plots using shannon index. There are also indices like simpson or invsimpson that one can calculate from the dataset.

## Shannon index

```{r}
diversity(BCI, index = 'shannon')
```

## Simpson index

```{r}
diversity(BCI, index = 'simpson')
```

## Inverse Simpson index

```{r}
diversity(BCI, index = 'invsimpson')
```

# Some plots comparing the indices

The indices are quite related to each other as indicated in the following plots.

Relationship between shannon and simpson indices of diversity.

```{r}
plot(diversity(BCI, index = 'shannon'), diversity(BCI, index = 'simpson'))
```

Relationship between shannon and invsimpson indices of diversity.

```{r}
plot(diversity(BCI, index = 'shannon'), diversity(BCI, index = 'invsimpson'))
```

Relationship between simpson and invsimpson indices of diversity.

```{r}
plot(diversity(BCI, index = 'simpson'), diversity(BCI, index = 'invsimpson')) # curvilinear relationship
```

## Fisher alpha index

We can also calculate fisher.alpha for the dataset using the function `fisher.alpha()`.

```{r}
fisher.alpha(BCI)
```

# Species cummulative curve

Further, we can plot species accumulation curve for the dataset. This is showing how the number of species vary with number of plots sampled. Normally the curve flattens at some point in which case additional sampling does not yield any more new species.

```{r}
plot(specaccum(comm = BCI, method = 'random', permutations = 100))

# other methods rarefaction, site, collector, exact, coleman
```

# Detrended correspondence analysis

Lastly, we can check for **detrended** correspondence analysis by plot using `decorana()` function.

```{r}
plot(decorana(BCI))
```

# Working with mite data

This has two separate data-frames, one for species per plot and the other for the environmental variables recorded from the same sampling plots as the species.

```{r}
data(mite)
data("mite.env")
```

```{r}
summary(mite.env)
```

```{r}
mite_hel <- decostand(mite, method = 'hellinger')
nmds1 <- metaMDS(mite_hel, autotransform = F)
```

```{r}
ordiplot(nmds1)
```

```{r}
ordiplot(nmds1, type = 't')
```

```{r}
autoplot(nmds1)
```

Here I make some informative plots.

```{r}
fort <- fortify(nmds1)
ggplot() +
  geom_point(data = subset(fort, Score == 'sites'),
             mapping = aes(x = NMDS1, y = NMDS2),
             color = 'black',
             alpha = 0.5) +
  geom_segment(data = subset(fort, Score == 'species'),
               mapping = aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
               arrow = arrow(length = unit(0.015, 'npc'),
                             type = 'closed'),
               color = 'darkgray',
               size = 0.8) +
  geom_text(data = subset(fort, Score == 'species'),
            mapping = aes(label = Label, x = NMDS1 * 1.1, y = NMDS2 * 1.1)) +
  geom_abline(intercept = 0, slope = 0, linetype = 'dashed', size = 0.8, color = 'gray') +
  geom_vline(aes(xintercept = 0), linetype = 'dashed', size = 0.8, color = 'gray') +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = 'black'))
```

```{r}
p1 <- ggplot() +
  geom_point(data = subset(fort, Score == 'sites'),
             mapping = aes(x = NMDS1, y = NMDS2),
             color = 'black',
             alpha = 0.5) +
  geom_segment(data = subset(fort, Score == 'species'),
               mapping = aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
               arrow = arrow(length = unit(0.015, 'npc'),
                             type = 'closed'),
               color = 'darkgray',
               size = 0,
               alpha = 0) +
  geom_text(data = subset(fort, Score == 'species'),
            mapping = aes(label = Label, x = NMDS1 * 1.1, y = NMDS2 * 1.1),
            alpha = 0) +
  geom_abline(intercept = 0, slope = 0, linetype = 'dashed', size = 0.8, color = 'gray') +
  geom_vline(aes(xintercept = 0), linetype = 'dashed', size = 0.8, color = 'gray') +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = 'black'))
p1
```

```{r}
p2 <- ggplot() +
  geom_point(data = subset(fort, Score == 'sites'),
             mapping = aes(x = NMDS1, y = NMDS2),
             color = 'black',
             alpha = 0) +
  geom_segment(data = subset(fort, Score == 'species'),
               mapping = aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
               arrow = arrow(length = unit(0.015, 'npc'),
                             type = 'closed'),
               color = 'darkgray',
               size = 0.8) +
  geom_text(data = subset(fort, Score == 'species'),
            mapping = aes(label = Label, x = NMDS1 * 1.1, y = NMDS2 * 1.1)) +
  geom_abline(intercept = 0, slope = 0, linetype = 'dashed', size = 0.8, color = 'gray') +
  geom_vline(aes(xintercept = 0), linetype = 'dashed', size = 0.8, color = 'gray') +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = 'black'))
p2
```

```{r}
ggarrange(p1, p2, ncol = 1)
```

```{r}
adonis(mite ~ Shrub, data = mite.env)
```

```{r}
p3 <- ggplot() +
  geom_point(data = subset(fort, Score == 'sites'),
             mapping = aes(x = NMDS1, y = NMDS2, color = mite.env$Shrub),
             alpha = 0.5) +
  geom_segment(data = subset(fort, Score == 'species'),
               mapping = aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
               arrow = arrow(length = unit(0.015, 'npc'),
                             type = 'closed'),
               color = 'darkgray',
               size = 0,
               alpha = 0) +
  geom_text(data = subset(fort, Score == 'species'),
            mapping = aes(label = Label, x = NMDS1 * 1.1, y = NMDS2 * 1.1),
            alpha = 0) +
  geom_abline(intercept = 0, slope = 0, linetype = 'dashed', size = 0.8, color = 'gray') +
  geom_vline(aes(xintercept = 0), linetype = 'dashed', size = 0.8, color = 'gray') +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = 'black'),
        legend.position = c(0.8, 0.2)) +
  scale_color_discrete("Shrubs")
p3
```

```{r}
ggarrange(p3, p2, ncol = 1)
```

# References
