---
title: "**Introduction to vegan package in R**"
author: "Wyclife Agumba Oluoch (wyclifeoluoch@gmail.com) </cent"
date: "<center>`r Sys.time()`</center>"
bibliography: 
  - bib/packages.bib
nocite: '@*'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: darkly
    highlight: pygments
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r libs, echo = FALSE, warning = FALSE, message = FALSE, include = FALSE}
packages <- c("base",
              'knitr',
              'rmarkdown',
              'prettydoc',
              'vegan',
              'ggvegan',
              'ggpubr',
              "tidyverse")
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}
lapply(packages, library, character.only = TRUE) |> invisible()
```

```{r write_bib, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
knitr::write_bib(c(
  .packages(), packages
), 'bib/packages.bib')
```

# Background----

In this write-up I explore `vegan` package. `vegan` is an `R` library that features a number of ecology relevant functions. Here I demonstrate some use cases.

# Load data----

Load data that come with the `vegan` package itself.

```{r}
data("BCI")
```

There are plenty data, we can check the dimensions of the data frame using the `dim()` function from base `R` package.

```{r}
dim(BCI)
```

The other way of getting the number of species per sampling plot in the data-set is to use `specnumber()` function within the `vegan` package.

```{r}
specnumber(BCI)
```

We can have the `specnumber(BCI)` converted into df and arranged to show which sampling plot had the highest number of species and which had the least.

```{r}
specnumber(BCI) |> 
  as.data.frame() |> 
  arrange(-specnumber(BCI))
```

It is possible to estimate the number of species in the pool using `specpool()` function from vegan in `R`. There are 225 species, but the estimator gives values from 236 to 247 with some standard errors around them. Would be interesting to go to the back-end and look at what goes on with this function.

```{r}
specpool(BCI, 
         smallsample = TRUE)
```

# Calculating some diversity indices

Here we explore a data-set comprising 50 field plots indicating the number of individuals of species of trees counted in each sampling plot. Typical ecological survey work.

We can calculate species diversity in the 50 plots using **shannon** index. There are also indices like **simpson** or **invsimpson** that one can calculate from the data-set.

## Shannon index

```{r}
diversity(BCI, 
          index = 'shannon') # Just specify the index to be shannon
```

## Simpson index

```{r}
diversity(BCI, 
          index = 'simpson') # Changing the code to calculate the simpson diversity index
```

## Inverse Simpson index

```{r}
diversity(BCI, 
          index = 'invsimpson') # The code to run inverse of simpson index
```

# Plots that compare the indices

The indices are quite related to each other as indicated in the following plots.

Relationship between Shannon and Simpson diversity indices.

```{r}
plot(diversity(BCI, 
               index = 'shannon'), 
     diversity(BCI, 
               index = 'simpson')) # Kind of linear
```

Assessing the relationship between shannon and invsimpson indices of diversity.

```{r}
plot(diversity(BCI, 
               index = 'shannon'), 
     diversity(BCI, 
               index = 'invsimpson'))
```

Relationship between `simpson` and `invsimpson` indices of diversity.

```{r}
plot(diversity(BCI, 
               index = 'simpson'), 
     diversity(BCI, 
               index = 'invsimpson')) # curvilinear relationship
```

## Fisher alpha index

We can also calculate fisher.alpha for the data-set using the function `fisher.alpha()`.

```{r}
fisher.alpha(BCI)
```

# Species cumulative curve

Further, we can plot species accumulation curve for the data-set. This is showing how the number of species vary with number of plots sampled. Normally the curve flattens at some point in which case additional sampling does not yield any more new species.

```{r}
plot(specaccum(comm = BCI, 
               method = 'random', 
               permutations = 100))

# Other methods include rarefaction, site, collector, exact, coleman
```

# Detrended correspondence analysis

Lastly, we can check for **detrended** correspondence analysis by plot using `decorana()` function.

```{r}
plot(decorana(BCI))
```

# Engaging mite data

This has two separate data-frames, one for species per plot and the other for the environmental variables recorded from the same sampling plots as the species.

```{r}
data(mite)
data("mite.env")
```

```{r}
summary(mite.env)
```

Calculating the `metaMDS` on the data. It is important to look into what these functions do and how to interpret their results

Calculating `ordiplot`. This is another plot that may need deeper understanding on how to interpret it and what it means.

```{r}
ordiplot(nmds1)
```

```{r}
ordiplot(nmds1, 
         type = 't')
```

```{r}
autoplot(nmds1)
```

Here I make some informative plots with forti.

```{r}
fort <- fortify(nmds1)
ggplot() +
  geom_point(data = subset(fort, 
                           Score == 'sites'),
             mapping = aes(x = NMDS1, 
                           y = NMDS2),
             color = 'black',
             alpha = 0.5) +
  geom_segment(data = subset(fort, 
                             Score == 'species'),
               mapping = aes(x = 0, 
                             y = 0, 
                             xend = NMDS1, 
                             yend = NMDS2),
               arrow = arrow(length = unit(0.015, 
                                           'npc'),
                             type = 'closed'),
               color = 'darkgray',
               size = 0.8) +
  geom_text(data = subset(fort, 
                          Score == 'species'),
            mapping = aes(label = Label, 
                          x = NMDS1 * 1.1, 
                          y = NMDS2 * 1.1)) +
  geom_abline(intercept = 0, 
              slope = 0, 
              linetype = 'dashed', 
              size = 0.8, 
              color = 'gray') +
  geom_vline(aes(xintercept = 0), 
             linetype = 'dashed', 
             size = 0.8, 
             color = 'gray') +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = 'black'))
```

```{r}
p1 <- ggplot() +
  geom_point(data = subset(fort, Score == 'sites'),
             mapping = aes(x = NMDS1, y = NMDS2),
             color = 'black',
             alpha = 0.5) +
  geom_segment(data = subset(fort, Score == 'species'),
               mapping = aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
               arrow = arrow(length = unit(0.015, 'npc'),
                             type = 'closed'),
               color = 'darkgray',
               size = 0,
               alpha = 0) +
  geom_text(data = subset(fort, Score == 'species'),
            mapping = aes(label = Label, x = NMDS1 * 1.1, y = NMDS2 * 1.1),
            alpha = 0) +
  geom_abline(intercept = 0, slope = 0, linetype = 'dashed', size = 0.8, color = 'gray') +
  geom_vline(aes(xintercept = 0), linetype = 'dashed', size = 0.8, color = 'gray') +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = 'black'))
p1
```

```{r}
p2 <- ggplot() +
  geom_point(data = subset(fort, Score == 'sites'),
             mapping = aes(x = NMDS1, y = NMDS2),
             color = 'black',
             alpha = 0) +
  geom_segment(data = subset(fort, Score == 'species'),
               mapping = aes(x = 0, 
                             y = 0, 
                             xend = NMDS1, 
                             yend = NMDS2),
               arrow = arrow(length = unit(0.015, 'npc'),
                             type = 'closed'),
               color = 'darkgray',
               size = 0.8) +
  geom_text(data = subset(fort, Score == 'species'),
            mapping = aes(label = Label, x = NMDS1 * 1.1, y = NMDS2 * 1.1)) +
  geom_abline(intercept = 0, slope = 0, linetype = 'dashed', size = 0.8, color = 'gray') +
  geom_vline(aes(xintercept = 0), linetype = 'dashed', size = 0.8, color = 'gray') +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = 'black'))
p2
```

```{r}
ggarrange(p1, p2, ncol = 1)
```

```{r}
#adonis(mite ~ Shrub, data = mite.env)
```

```{r}
p3 <- ggplot() +
  geom_point(data = subset(fort, 
                           Score == 'sites'),
             mapping = aes(x = NMDS1, 
                           y = NMDS2, 
                           color = mite.env$Shrub),
             alpha = 0.5) +
  geom_segment(data = subset(fort, 
                             Score == 'species'),
               mapping = aes(x = 0, 
                             y = 0, 
                             xend = NMDS1, 
                             yend = NMDS2),
               arrow = arrow(length = unit(0.015, 'npc'),
                             type = 'closed'),
               color = 'darkgray',
               size = 0,
               alpha = 0) +
  geom_text(data = subset(fort, 
                          Score == 'species'),
            mapping = aes(label = Label, 
                          x = NMDS1 * 1.1, 
                          y = NMDS2 * 1.1),
            alpha = 0) +
  geom_abline(intercept = 0, 
              slope = 0, 
              linetype = 'dashed', 
              size = 0.8, 
              color = 'gray') +
  geom_vline(aes(xintercept = 0), 
             linetype = 'dashed', 
             size = 0.8, 
             color = 'gray') +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(color = 'black'),
        legend.position = c(0.8, 0.2)) +
  scale_color_discrete("Shrubs")
p3
```

```{r}
ggarrange(p3, 
          p2, 
          ncol = 1)
```

# References
